---
title: "Best Model Selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{best-model-selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `Best_Model` flag identifies the selected forecast for each series. Finn first checks candidate validity, then balances backtest weighted MAPE with future-forecast plausibility. Individual models, learned ensembles, and simple averages are evaluated on the same series and expected backtest dates.

## Process at a glance

Model selection is several connected decisions, not one minimum-error calculation:

| Stage | Question answered | Result |
| --- | --- | --- |
| Candidate validity | Does each model have complete, finite predictions and usable backtest accuracy? | Hard-invalid candidates cannot win or contribute to a simple average. |
| Within-run selection | Among sufficiently accurate candidates, which future path ranks best by risk, concern count, supported seasonal fidelity, WMAPE, and stable ID? | One selected model or exact average for each series. Soft concerns can remain when no better eligible alternative exists. |
| Agent iteration context | Which completed iteration's settings should guide further optimization? | A near-best iteration can be preferred when its individual model-pool accuracy improves. This does not automatically replace a better saved local forecast. |
| Saved forecast promotion | Should the current iteration replace the saved result? | Local winners stay protected; globally selected series reference one winning global iteration. |
| Update acceptance | Is the newly refitted version of the saved choice still acceptable? | Keep that choice or attempt the existing default-local replacement workflow before reconciliation. |

Seasonal fidelity breaks a risk-and-concern tie only when every candidate in that
tied group has a finite assessed score; otherwise WMAPE and stable ID decide the tie.

For a standard run, follow [candidate checks](#future-forecast-checks), the
[accuracy allowance](#accuracy-allowance), and
[selection and saved averages](#standard-forecasts-and-saved-averages).
The [Agent](#iterative-agent-forecasts) and [update](#updated-agent-forecasts)
sections describe the additional decisions. [Reconciliation](#reconciliation)
makes the selected hierarchy coherent; it does not select the models again.

Two uses of "average" are important: a **simple-average forecast** combines
predictions from selected components, while **average model accuracy** summarizes
individual model WMAPEs to guide Agent search. Improving the latter can be useful
even when the winning model has not changed.

## Experimental custom models

`prep_models()` and `forecast_time_series()` accept a named `custom_models` list.
Only names explicitly included in `models_to_run` become candidates. Supplying
custom models alone does not change the built-in default pool; an explicit pool
can contain custom models only or a mixture of custom and built-in models for
`bottoms_up`. Custom hierarchy execution requires an explicitly selected mixed
pool and separately reviewed hierarchy capabilities.

`create_custom_model()` is the single natural-language authoring entrypoint.
It accepts raw `input_data` with Finn metadata or an existing prepared `run_info`.
Humans and coding agents use the same function, including continuation through
`draft` and `responses`. Advanced settings belong in `custom_model_control()`;
approval stays visible at the top level and defaults to manual.

New calls default to local models. Raw input must contain a column named `Date`
and an explicit target and horizon. Grouping can be omitted only when dates are
unique and no varying undeclared columns suggest multiple series. Cadence can be
omitted only when at least three actual dates per series match exactly one
complete day/week/month/quarter/year grid. Ambiguous or incomplete grids require
explicit metadata. Prepared runs supply their saved settings, including forecast
approach. Neither inference supplies fiscal calendars nor proves business close.

Raw driver-based models can be created and validated before the caller creates
its forecast run or calls `prep_data()`. Authoring keeps original driver names,
values and types alongside its privately prepared historical target data.
Only observed historical driver rows enter validation; future values are not
used, and missing/nonfinite driver history is rejected rather than filled.
Check forecast-date coverage and supply the required future driver inputs before
executing the resulting model. Private temporary preparation still occurs
inside authoring; it does not create the caller's forecast run.

Use `hist_end_date` as the last closed actual period, with the same date-label
convention as the data. For monthly observations labelled on the first day,
August 2026 is `as.Date("2026-08-01")`, not August 31. Without an explicit
cutoff, every series must have the same last finite target date for inference.
Finite values do not prove a business close: supply a cutoff when later target
rows contain budgets, prior forecasts or partial actuals.

Before sampling or padding, raw input must cover an uninterrupted common
historical window with finite targets for every series. Drivers must cover that
window and every requested future period; future target values are ignored.
Missing periods, uneven histories and future-only series now fail rather than
being padded with zeros. Explicit common `hist_start_date`/`hist_end_date`
bounds can select a complete overlap. Prepared driver artifacts are checked
across all requested series before sampling, but their stored values cannot
prove whether earlier preparation filled a raw gap.

The authoring cutoff is saved validation evidence, **not a permanent model
cutoff**. An approved definition can be refitted in later runs with different
history end dates and new driver values, as long as its declared capabilities
remain compatible. Each fit derives its own cutoff and per-series cutoffs;
relative calculations must use those dates and the prediction request. Reusing
an old fitted workflow without refitting does not advance its cutoff. A pending
authoring draft, unlike a reusable approved model, retains its original inputs.

Driver-based holdout metrics are explicitly conditional on supplied assessment
drivers. The current interface does not verify historical plan vintages or
publication delays. For realistic business backtests, provide the plans actually
available at each forecast origin rather than revised plans or realized drivers.

Enrollment accepts only `finnts_custom_model` envelopes
containing an exact passive definition, version-bound technical check summaries,
manual intent confirmation or explicit automatic trusted-code authorization. Raw definitions, registry
references, fitted models and stale approval evidence are rejected. Attestations
are provided by a trusted caller; they are not authenticated proof of human
approval or a sandbox for untrusted R code. Tests use clearly synthetic evidence,
not authenticated proof of approval or live-provider validation.

The default `approval = "manual"` asks unresolved business questions, prepares a fixed
interpretation and numerical examples, and generates a candidate. One compact
review then asks whether to create that exact model. A yes both authorizes
same-session execution and conditionally approves the model if validation passes.
There is no second approval prompt after success. A repair cannot change the
expected values or reuse another version's consent: revised code always receives
a fresh review. Holdout accuracy is reported, not used to rewrite the human rule.

If validation fails and another candidate is proposed, manual review now explains
why approval is needed again. It displays the prior failure phase and safe reason,
previous/current version IDs, and the names of changed source functions. The rule
and worked examples remain unchanged; the proposed repair is not yet validated.
An identical resubmission is labelled as the same failed candidate, not a fix.
`details` shows full identities and the current source without executing it.
Compact per-function hashes support the comparison without retaining another
full source snapshot. Older diagnostics can lack those hashes, in which case
comparison detail is explicitly unavailable. Existing saved reviews remain valid;
exact-candidate consent, replay handling and retry limits do not change.

```{r custom-authoring-example, eval=FALSE}
llm <- ellmer::chat_openai()
history <- data.frame(Date = as.Date(c("2020-01-01", "2020-02-01")), Target = c(10, 20))
future <- data.frame(Date = seq(as.Date("2020-03-01"), by = "month", length.out = 3))
examples <- list(
  custom_model_example(history, future, expected = c(15, 15, 15)),
  custom_model_example(transform(history, Target = 0), future,
    expected = c(0, 0, 0), edge_case = TRUE)
)
custom <- create_custom_model(
  instructions = "Use the mean of all historical targets independently for each series.",
  llm = llm,
  input_data = data_tbl,
  combo_variables = "id",
  target_variable = "value",
  date_type = "month",
  forecast_horizon = 3,
  name = "business_mean",
  validation_examples = examples
)
```

`custom_model_example()` accepts history, future rows and independently calculated
answers. Numeric answers bind to the original future row order before sorting;
there is no recycling or inferred answer. `target_variable` maps a caller target
to `Target`, and `combo_variables` maps declared series keys to `Combo`. Omitted
keys create one example series; global examples require explicit series coverage.
For a deliberate error case, supply `expected_error = list(phase = "fit", code =
"zero_denominator")` and omit `expected`. At least two cases, every declared
mode, and an edge case remain required by creation. Existing example lists still
work; every supplied expectation remains mandatory during repair.

`custom_model_control()` has seven settings: `max_attempts = 3L`,
`validation_budget = 60`, `package_policy = "installed_finnts"`,
`required_properties = character()`, `forecast_approach = NULL`,
`draft_path = NULL`, and `interact = NULL`. Build a fresh control to change
settings; explicit-setting markers distinguish supplied values from defaults.
The public signature exposes these settings only through `control`. Supply
`approval`, `control`, `draft` and `responses` by name. Unknown arguments fail;
omitted controls never overwrite saved draft settings. `validation_budget` is
not a hard timeout. `control$draft_path` controls persistence only, including
synchronous callback or console review. Custom-model creation supports one
current implementation; obsolete development arguments and artifacts must be
recreated, not migrated through compatibility bridges.

In an interactive R session, answer `y`/`yes` or `n`/`no`, without typing hashes.
Case and surrounding whitespace are ignored; blank or EOF declines. `details`
shows readable source and complete examples without executing or regenerating
anything. `edit` collects a requested change and prepares a new review within
the cumulative proposal limits. Existing data/capability constraints stay fixed.
The default display omits instructions, schema/package inventories and source
dumps; a short measured summary follows success.

A trusted `control$interact` callback receives `stage`, `prompt`, `identity`, `review`
and `questions`. Stage `clarify` requires named text answers. Stage `review_create`
accepts the user's yes/no decision, `details`, or
`list(action = "edit", instructions = "requested change")`. Present the review
to a user; never implement unconditional approval. Stale or exhausted responses return no
enrollable model. See `?create_custom_model` for the complete contract.

### Deterministic Authoring Mechanics

The current authoring contract uses typed structured responses. The default
`control$package_policy = "installed_finnts"` freezes installed direct FinnTS `Depends`,
`Imports` and `Suggests`, plus the existing R base/recommended allowance. It does
not admit arbitrary installed or transitive-only packages. Use
`custom_model_control(package_policy = "base_only")` to restrict generated code to base R.

FinnTS parses source to derive only the packages actually referenced through
`pkg::function`; the LLM no longer supplies a dependency list. The resulting
packages and complete assembled source are hashed before candidate review and
authorization. Nothing is installed or attached. Namespace availability and
public exports are checked after authorization, before candidate bodies execute.
Namespace loading can itself have side effects; it is not a sandbox.

FinnTS also supplies fit/predict signatures, prediction row IDs, and generated
example dates, keys, declared-mode coverage and tolerance. The LLM returns
calculation bodies, optional helpers, synthetic value arrays, independent expected
answers and business rationale. Prediction bodies return finite numeric vectors
in request order; the fixed wrapper rejects wrong lengths/types rather than
recycling, clipping or imputing. The full reusable model remains ordinary R.

Authoring metadata lists only the workflow's actual input columns:
`Date`, `Combo`, `Target` and declared external regressors. Prediction excludes
`Target`. Internal prepared features such as `Date_index.num` are not advertised
as implicit inputs; an explicitly declared driver with that name remains valid.
For a six-observation trend, generated code can construct positions `1:6` and
predict at `6 + context$forecast_step`, without assuming a prepared feature exists.
Only the outgoing metadata copy is filtered. Captured data, saved metadata,
accepted examples, approved model identities and older authoring protocols stay
unchanged.

For new definitions, FinnTS supplies fit and prediction rows sorted by series
and Date, with integer `context$forecast_step` values derived from each series'
current training cutoff. It restores the caller's original output order.
Each included series must supply its complete forecast horizon; partial-date
requests fail before prediction source executes. The runtime does not synthesize
missing drivers or infer fiscal calendars. Daily through yearly cadences use
their saved period labels. Separate series batches are allowed when the business
rule permits them, not as a guarantee that pooled rules are independent.
Later refits derive new cutoffs without changing model identity. Saved older
definitions keep their original ordering and partial-request behavior.
Internal resampling may have shorter scoring windows. FinnTS supplies existing
predictor rows through the full horizon, then retains only the original scoring
rows. It does not extend training history, expose assessment targets to generated
prediction code, fill missing drivers, or change forecast-selection policy.

Body fields prefer calculation statements, but new proposals also accept one
complete function with the exact signature `function(data, context, parameters)`
for fit or `function(object, new_data, context)` for prediction. No defaults,
ellipsis, reordered names or extra arguments are allowed in this alternate form.
FinnTS extracts that function body once using parsed syntax, before hashing,
review and authorization. It never calls the function to discover its result.
Equivalent complete-function and body inputs yield the same canonical candidate.

Ambiguous signatures, extra top-level expressions alongside a function, and
direct nested function returns produce `source_body_format` feedback before
execution. Local assigned helper functions and callbacks remain valid. Fitted
state must still be portable data and prediction must still return a numeric
vector. Saved-source replay does not normalize or silently repair old models;
their exact source, identity and authorization remain unchanged. This support
does not introduce another authoring protocol or any new public argument.

Generated examples are native nested objects and arrays in the structured
response, not JSON embedded inside a text field. `series` remains an array even
for one series. FinnTS validates required fields, lengths and captured numeric,
character/factor or logical predictor types locally as well. The missing-data
policy is fixed to rejection of required missing inputs; the LLM cannot return
a blank replacement. This adds neither cleaning nor imputation.

The arbitrary passive fixed-parameter map uses `parameters_json` to support
structured parameter values. Invalid JSON identifies that field explicitly;
malformed strings are not repaired by guessing braces or stripping text. Expected answers remain
independent of candidate execution. Structural validation does not independently
prove their arithmetic. New calls treat LLM-generated numerical expectations as
diagnostics: mismatches produce a user warning, not rejection or a request to
rewrite the formula. Measured errors remain in the validation evidence, and the
warning is retained in `model$validation$checks$llm_example_warning`. Matching
incorrect code and answers can still pass finite comparisons; a warning alone
cannot identify which side is wrong. Current-format drafts retain their saved policy.

**Bring your own validation examples.** Calculate expected outputs with business
logic reviewed independently of the generated model, then pass the materialized
cases through `validation_examples`. This is an examples-list argument, not a
validation-function callback. Caller-provided output mismatches remain mandatory
and can trigger bounded source repair. These exact examples are included in the
initial contract request, so the provider can align history requirements and
preserve caller error codes/phases before generating source. It must not invent
extra business error declarations without matching caller error cases.
Include at least two cases, every declared
mode, and meaningful edges such as zeros, negative values or calendar boundaries.
For example, these independently specified checks describe a three-point mean:

```{r caller-validation-example, eval=FALSE}
history <- data.frame(Date = as.Date(c("2020-01-01", "2020-02-01", "2020-03-01")),
  Combo = "example", Target = c(10, 20, 30))
future <- data.frame(Date = as.Date(c("2020-04-01", "2020-05-01")), Combo = "example")
ordinary <- list(name = "mean", model_type = "local", history = history,
  new_data = future, expected = transform(future, .pred = 20), tolerance = 1e-8,
  edge_case = FALSE, rationale = "(10 + 20 + 30) / 3 = 20")
zero <- ordinary
zero$name <- "zero"
zero$history$Target <- 0
zero$expected$.pred <- 0
zero$edge_case <- TRUE
zero$rationale <- "The mean of three zeros is zero."
validation_examples <- list(ordinary, zero)
```

Use those cases only for that rule and matching cadence/horizon, not for arbitrary
custom logic. Caller examples are also provider-visible; keep sensitive real
observations out of examples you do not want sent to the provider. Technical
failures such as broken R, invalid output shape, nonportable state, required
properties and coded-error mismatches still fail and may use bounded repair.
Only caller-supplied numerical expectations can request an output-matching
rewrite. LLM-generated numerical differences cannot.

Parameters have explicit types and units. Declared numeric weight or
decay vectors become atomic R vectors; objects and arrays of records retain list
structure. Synthetic numeric history and future columns share double types,
including whole-valued historical discounts followed by fractional forecasts.
Blank nonnumeric units normalize to `not_applicable`.
Numeric units remain required; no business units are inferred. Malformed
declarations receive bounded field-specific diagnostics before contract acceptance.

Required business errors are separate outcomes, not placeholder zero forecasts.
An `error_policy` binds the code and fit/predict stage, and each mode must include
valid numerical controls plus tests for its declared errors. Generated error
cases use empty expected arrays. Caller error cases use `expected = NULL`,
`outcome = "error"`, and `expected_error = list(phase = "predict",
code = "zero_denominator")` with the usual example metadata and nonempty tables.
Candidate code raises the portable sibling helper `finntsRuleError(code)`.
An unrelated exception or a zero-returning fallback does not pass that test.
Additional named error scenarios remain within the existing eight-case limit.

Row-order invariance and fitted-state serialization are mandatory. Other properties
are tested when declared: constant forecasts, target-scale equivariance,
independent series, fixed history windows, and relative calendars. The
caller `control$required_properties` are frozen hard requirements. Other LLM suggestions
are diagnostic only: their failed checks are reported without rejecting valid
calculations. This does not weaken caller numerical cases, coded business errors, row identity, finite outputs,
portability or exact-version approval. For example, use
`custom_model_control(required_properties = "constant_forecast")` only when constant output is a genuine
business requirement, not for every custom model. Calendar tests
translate inputs by four years; fixed-calendar rules should not declare that
property. Reports retain measured counts and expectation provenance, explicitly
labelled caller-supplied or LLM-proposed and unverified. These are behavioral
checks, not a general rule compiler or independent arithmetic verifier.

Stateless future-driver formulas may declare zero formula-history need. Their
workflow examples still need a historical row to establish the schema and cutoff;
FinnTS records that workflow minimum separately. Preprocessing casting failures
stop rather than requesting different source arithmetic. Repeating the same
source version and failure stops the repair loop without changing fixed answers.
Contradictory contracts require a new authoring version and authorization.
New source responses may explicitly report `contract_conflict`; this stops before
execution without rewriting the accepted contract or answers. Under the new
comparison policy, disagreement with an LLM-generated numerical answer is not
a reason to declare a conflict: source should implement the business rule and
let validation record any numerical warning. Tagged protocol
questions receive package-owned facts within the existing retry budget, while
unresolved business questions still require clarification. Automatic defaults
cannot override explicitly supplied settings. Invalid defaults declarations are
corrected as contract defects, not by changing the caller's settings.
Current-format approved models are not silently revalidated or repaired.

Generated examples use one basic and one edge scenario per declared mode, with a
fixed cadence-aligned synthetic cutoff and enough provider-supplied historical
values for the rule. Caller-provided examples retain their original tables,
values and tolerances. No candidate execution supplies its own expected answers.
Those answers and business interpretation remain fallible LLM work: this is not
an independent general formula evaluator. New automatic calls supply a
percentage-growth convention through the current defaults catalog, but that policy does
not prove arbitrary provider prose or code follows it.

The contract requests explicit `history_requirements`: a nonnegative integer
`minimum_rows_per_series` and unique positive `required_lag_periods`, bounded
by the existing example limits. Offsets use the captured cadence. Rules without
exact observed-history lookups may declare an empty lag array. FinnTS validates
each example series against the minimum count and every required lag date before
accepting the contract. For monthly same-month values and three annual
differences, offsets 12, 24, 36 and 48 require four annual observations; seven
monthly rows cannot meet that requirement.

Fit-time windows use the minimum count, not forecast-relative lag offsets.
For example, "use the median of the last six actual months for every forecast
month" needs `minimum_rows_per_series = 6` and `required_lag_periods = []`.
The model computes that median at fit time and repeats it across the horizon.
Declaring lags 1 through 6 instead asks for observations before each future
forecast date. With three forecast months, offsets 1 and 2 eventually point
into the unobserved forecast window. More old observations cannot fix this.

Repair feedback explicitly identifies `lag_reference = "forecast_date"` and
the conflicting `unobserved_lag_periods`. Exhaustion errors include the same
bounded explanation. A provider must correct an unaccepted declaration that
misrepresents the rule; FinnTS never silently drops its lags or changes expected
answers. Seasonal forecast-date lookups retain their existing strict checks.

Missing coverage is rejected during contract review, before source generation.
Authoring can retry an unaccepted history declaration that conflicts with
structurally valid caller examples. For example, a fit-time growth window must
not be described as observed lags reaching into the forecast window. The next
attempt receives numerical-case row counts and impossible-lag feedback;
intentional error cases are excluded from those counts. The declaration must
still express the actual business rule, not weakened history needs merely to
pass. Exhaustion returns an error under the existing contract-attempt limit.
Only unaccepted generated examples may be replaced. Caller examples, accepted answers
and the source-repair oracle never change, and no history is invented, imputed or
silently extended. The declaration is testable metadata, not a proof that the LLM
has interpreted every possible business rule correctly.

New source preflight also rejects obvious uninitialized protocol-state references,
such as `object$history <- data` in fit when `object` was never created. Fit receives
`data`, `context` and `parameters`; prediction receives `object`, `new_data` and
`context`. The check never inserts an object or rewrites a formula. It inspects
ordinary evaluated syntax and lexical bindings without execution, while leaving
quoted/formula and unknown nonstandard-evaluation syntax alone. Uncertain control
flow still requires runtime validation. Bounded `source_binding` evidence names
the function and symbol for source repair, without arbitrary exception text.

Drafts and approved models never gain new permissions on resume or forecast reuse.
Omit `control$package_policy` when
resuming; an explicit policy change requires a new call. Unknown protocols,
altered scaffolds and changed runtime/package metadata are rejected.

### Automatic Authorization

Set `approval = "automatic"` to authorize generated R and bounded repairs without
manual review. It never calls `readline()`, requests clarification or returns a
human-review draft: it returns a validated model or a typed actionable error.
`control$interact` and `responses` are prohibited. This deliberately gives up source
review, not numerical validation. The code runs in the current R session with
your permissions, without a sandbox or reliable interruption.

```{r custom-authoring-automatic, eval=FALSE}
custom <- create_custom_model(
  instructions = "Use the mean of historical targets independently for each series.",
  llm = llm, input_data = data_tbl,
  combo_variables = "id", target_variable = "value",
  date_type = "month", forecast_horizon = 3,
  approval = "automatic"
)
```

Automatic calls use one closed, package-owned defaults catalog. Explicit supported
instructions and supplied metadata take precedence; incompatible or contradictory
requirements fail instead of silently widening capabilities.

| Omitted Choice | Default or Required Action |
| --- | --- |
| Target and horizon | Required for raw input; inherited from compatible prepared metadata. Never guessed. |
| Series keys and cadence | Only unambiguous single-series grouping and complete regular grids can be inferred; otherwise supply them explicitly. |
| Forecast approach | `bottoms_up` for raw input; inherited for prepared runs. Hierarchy retains capability restrictions and a mixed execution pool. |
| Model mode | `local`; global/pooling requires explicit `model_type` and compatible metadata. |
| Lookback | All training history within supplied historical bounds and the current fold cutoff. Explicit supported lookback wins. |
| Ordinary arithmetic-average weights | Equal weights only when unspecified. A requested weighted rule with missing weights fails. |
| Growth basis | Percentage change, `newer / older - 1`, when unspecified; apply the rate multiplicatively. Explicit absolute, percentage-point or CAGR rules override this fallback. |
| Drivers | None unless declared by arguments or prepared metadata. No automatic feature selection or derived columns. |
| Required history/predictors | Reject empty, missing or nonfinite values and duplicate series/date keys; no dropping or imputation. Unsupported cleaning fails. |
| Zero/negative values | Valid observations; no implicit clipping or rounding. |
| Undefined division | Error, with no invented epsilon, cap or rate. Zero-denominator holdout WMAPE remains unavailable. |
| Trend, seasonality, transforms, tuning | None implicitly added; use original-scale R1. |
| Dependencies | Freeze installed FinnTS eligibility, or explicit `base_only`; derive actual dependencies from syntax. No installation. |
| Worked examples | New calls scaffold basic/edge examples per mode; the provider supplies independent values/answers. Generated tolerance is `1e-8`. |
| Caller-provided examples | Preserve expected values and tolerances; never relax them during repair. |
| Attempt limit | Three attempts total per contract/source phase; `control$max_attempts` may be 1-3. |
| Validation budget | 60 seconds after evaluation returns; positive `control$validation_budget` may be at most 120. Not a hard interrupt. |
| Sampling | Up to three deterministic series/leaves and 10000 historical node rows. Raw hierarchy retains derived nodes; prepared hierarchy is read completely. |
| Name | Generate a safe descriptive name; preserve an explicit name. |

Never invent business offsets, multipliers, numeric growth rates, allocation percentages,
caps, commercial targets, required weights/drivers/formula terms, unspecified
fiscal calendars or aggregation semantics. If essential context remains unresolved,
start a new call with explicit instructions and compatible metadata. The catalog
does not infer these choices from observed values or improve the requested rule.

For a prior-year seasonal value plus average growth over three years, four
annual observations of 100, 120, 150 and 180 give percentage changes of 20%, 25%
and 20%. Their equally weighted arithmetic mean applied to 180 gives **219**.
This is not CAGR, an absolute increment or extra compounding across forecast
steps. The requested lookback and seasonal/calendar alignment still govern
which observations are used. Undefined division still errors; no epsilon,
clipping or fallback rate is introduced. Existing absolute-growth worked
examples and caller-supplied answers are never rewritten to fit this convention.

The fixed contract records `catalog_version`, effective settings and actually
applied defaults. The approved model retains their JSON in
`custom$validation$checks$automatic_defaults`; success prints the applied choices
and measured results. Semantic defaults-used declarations come from the provider
and are restricted to catalog keys/values. Neither those declarations nor finite
examples prove arbitrary prose/code semantics for every future input.
`growth_basis` is recorded only when the provider declares that the percentage
fallback was applied, not for explicit growth bases or non-growth rules.

Every source version receives a separate automatic authorization before execution.
The record has `intent_confirmed = FALSE`, `allow_code = TRUE` and
`mode = "automatic"`, bound to its `version_id`; it never claims human review.
Manual envelopes record the three explicit confirmation fields. Definition hashes
and `Model_ID` do not change solely because approval policy differs, while run and
Agent provenance distinguish the policies. Forecast reuse does not call the authoring LLM or regenerate source.

The only public approval values are `"manual"` and `"automatic"`. Explicit NULL,
`"interactive"` and `"deferred"` fail. Noninteractive manual calls return a review
draft. Omitted approval on resume inherits the explicitly saved mode; mode changes
fail. Automatic `control$draft_path` stores confidential
audit state without pausing; an approved current pointer replays without refitting.

### Example Dates and Retries

Every worked example must forecast exactly the next `forecast_horizon`
consecutive cadence dates immediately after its last historical Date. With
monthly horizon 3 and history ending December 2025, those dates are
January-March 2026. A prior-year seasonal rule uses January-March 2025 values
from history; it does not move the forecast window another year forward.
History must contain the observations the rule needs. For a fixed 11% increase,
prior-year values 10, 20 and 30 produce 11.1, 22.2 and 33.3. Zero and negative
values remain valid unless explicitly excluded by the supported rule.

Examples must cover every declared `model_type` and no undeclared modes. A
local-only contract needs only local examples, with exactly one series per
example; it must not add a global example merely because FinnTS supports global
models. A global-only contract needs only global examples, each with at least
two series. Either single-mode contract still needs two to eight total examples,
including an edge case. Mixed contracts require coverage of both declared modes.

An undeclared example mode produces specific mode feedback during contract
retries. FinnTS does not silently remove or relabel the example or enable another
mode to make it pass. The provider may regenerate the unaccepted proposal within
the existing attempt limit; caller-provided examples must instead be corrected
by the caller. Valid tolerance values are not reported as the cause of a
well-formed example's undeclared mode.

When a generated contract is rejected, the next bounded contract attempt receives
safe package-owned feedback explaining the rejection category and correction.
The final error includes the last safe rejection if attempts are exhausted.
Raw exception messages, private values, paths and provider output are not copied
into that feedback. A draft retains its diagnostic code and attempt counts.

Feedback also names the rejected field, such as `parameters_json` or
`examples.series`. Unexpected validator failures stop rather than consume more
provider retries. An optional `diagnostics$rejected_contract` snapshot retains
only known provider-visible response fields, their hash, the field/code and an
omission marker. It shares the existing one-MiB diagnostic cap; oversized content
is omitted without truncating an accepted oracle. Raw exceptions, Chat objects
and actual training data are excluded. This snapshot is diagnostic evidence,
not an accepted contract or authorization. No extra automatic file is written.

Automatic `business_clarification` errors retain the unanswered questions at
`diagnostics$rejected_contract$proposal$questions`, subject to the same size
limit. The normal error message also displays the questions, labelled as
provider clarification rather than independently verified input errors. Each
question is limited to 500 displayed characters with a truncation marker, and
control characters are removed; the original diagnostic text is unchanged.
No debugger-frame search is needed to see ordinary questions. Supplied metadata and catalog
defaults should resolve known choices; genuinely missing business terms still
require clarification rather than guessed answers.

For insufficient declared history, retries receive `feedback$history` with
package-computed minimum rows, the supplied row counts per scenario/series,
declared lags and whether all lookups can remain historical through the horizon.
Generated cases use scaffold identities; caller cases use their
already provider-visible names and series keys. Only counts and identifiers are
included in this feedback, not observed training values.
No rows or answers are filled or rewritten. New examples should use simple,
independently checkable arithmetic; source repairs cannot replace a bad accepted
oracle merely to pass validation.

The provider may replace invalid generated examples only before the contract is
accepted. Caller-provided examples are never replaced: structurally invalid
supplied examples fail when checked, without spending further provider attempts
on them. A conflicting unaccepted history declaration is retried
instead of blaming or changing a valid caller example; earlier protocols retain
their existing history-failure behavior. After
contract acceptance, code repairs must preserve its examples and tolerances.
The existing date checks, numerical checks, three-attempt limits and manual or
automatic authorization policy remain unchanged. Corrective feedback improves
repair information; it does not guarantee successful live generation.

### Source Repair Diagnostics

New source requests contain `fit_body`, `predict_body` and optional named helper
functions. Body fields may use the exact complete-function alternate form above.
FinnTS assembles the fixed interfaces and row-identity wrapper from
parsed expressions, without evaluating them. A passive preflight checks
syntax, permitted namespaces and resolvable base/lexical function calls, without
guessing packages for unresolved calls. Units come from the accepted
interpretation and examples: a lag of twelve monthly observations means one
year, not twelve years. Parameters needed by `predict()` belong in fitted state.

Dependency repairs also receive bounded export-location hints from already-loaded
eligible namespaces. FinnTS neither loads additional namespaces for this lookup
nor substitutes functions automatically; identical exported names may have
different semantics. Date lookup guidance preserves matching key classes:
`vapply()` can drop a `Date` result's class even with a Date template. These aids
reduce mechanical errors, but do not prove arbitrary generated logic correct.

Every custom model uses one current calendar-helper bundle. The reviewed
helper source includes:

- `finntsShiftDate(dates, periods, date_type, invalid_day = "error")` returns
  Date values in the supplied order for integral day, week, month, quarter or
  year offsets. Negative offsets look back; offsets must be scalar or match the
  dates' length. Nonexistent month-end or leap-day dates error by default.
  `invalid_day = "last_day"` is available only when the business rule explicitly
  permits clamping. The helper does not infer fiscal calendars.
- `finntsMatchDate(dates, history_dates, combo = NULL, history_combo = NULL)`
  returns historical row indices, preserving request order. Pooled histories
  must supply both character series-key vectors. Date vectors must retain their
  class, and historical keys must be unique within each series. Missing keys
  return NA rather than filling history or changing the model's error policy.
- `finntsRowsDate(dates, history_dates, history_combo)` returns
  a list of all matching historical row indices for each requested date.
  Repeated dates across different series are valid; duplicate series/date keys
  still error. Missing dates return `integer(0)`. The LLM implements the actual
  aggregation, weighting, grouping and business missing-history behavior.

These helpers handle calendar mechanics, not the business calculation. Their
source and the reserved `finntsDateError` helper are frozen into model identity,
repair snapshots and saved approvals. They require no new package dependency
or provider call during prediction and leave room for 24 provider helpers within
the unchanged source-size limit.

New global contracts distinguish shared fitting from prediction scope.
`prediction_cohort` declares `series`, `groups` with explicit stable group
columns, or `all` fitted series. A shared regression can predict complete series
batches independently; portfolio allocations generally need the full cohort.
FinnTS checks required peers and group membership before generated prediction
code runs. Resampling expands execution to required peers and dates while
retaining the original analysis rows and scoring only original assessment rows.
Local models use series scope; global models declare the scope their rule needs.

New explicit global authoring validates the complete supplied cohort under the
existing 10000 historical-row cap. Oversized inputs fail with subset guidance;
FinnTS does not silently drop peers or categories to fit the cap. Regression
guidance asks generated code to retain feature order, category levels, contrasts,
scaling and time origin in portable fitted state. These mechanics do not choose
formulas: the LLM still writes the actual model, transformations and allocation.

A genuine fixed-contract contradiction stops authoring. An explicitly reported
implementation difficulty can use only remaining source attempts; it cannot
change examples, extend the retry budget or bypass exact-version approval.
Provider failures are separate operational errors with bounded category, HTTP
status when known, phase, prompt bytes and elapsed time. Unknown causes remain
unknown; raw provider messages, response bodies and headers are not retained.

The shared source guard catches explicit Date-template `vapply()` and plainly
Date-returning `sapply()` simplification before execution. It preserves numeric
iteration, quoted code, local shadows and explicit Date/numeric conversions;
unknown expressions are not treated as proven type errors. Runtime helper
conditions distinguish `date_key_type`, `date_key_duplicate`, and
`date_shift_invalid` from genuine missing-history business errors. Together with
`date_class_loss`, these produce bounded technical repair feedback without
exposing lookup values, rewriting examples, increasing retries or bypassing
approval. They are not a general proof of generated-code correctness.

Unsupported development formats require recreation, not compatibility bridges.
Approved models never acquire helpers silently. Approved helper
source remains unchanged during later-cutoff refits, serialization and worker
execution, just like the model's calculation source.

Repairs receive the previous source and a version-bound failure record identifying
the phase, reason and, where applicable, worked-example index and mode. Fitting,
prediction, arithmetic mismatch, nonportable state, roundtrip disagreement,
holdout failure and elapsed-budget failure remain distinct. Up to four keyed
expected/actual comparisons may be provided for fixed worked examples. Real
holdout values, driver values and business-series identifiers are never added.

Recognized missing-object runtime errors now use `reason = "unbound_symbol"`
with a separate `failure$unbound_symbol` value, such as `history`. That value must
be an ASCII identifier of at most 120 bytes referenced in parsed candidate code.
Names found only in strings, member names, assignment bindings or quoted/formula
expressions do not qualify, and sensitive-looking identifiers are withheld.
Unrecognized or localized error wording remains generic. No raw exception text,
stack trace or runtime value is included in this diagnostic.

The source request instructs prediction code to bind fitted-state aliases
explicitly. For example, `history <- object$history` makes `history$Date` valid;
`with(object$history, ...)` alone exposes the table's columns but does not create
the `history` alias. Valid data masks and nested functions remain supported.
FinnTS reports runtime evidence rather than pretending to infer every R scope,
and never injects an alias or changes the generated calculation automatically.

Manual repair reviews and exhausted-attempt messages show the bounded missing
identifier. Repairs still preserve the accepted contract, caller examples and
attempt budget; manually repaired source still requires fresh version-specific
consent. Saved failures without the new field and existing approved models
remain readable without changing their source or triggering new authoring.

Messages are package-owned by default. An exact constant literal `stop()` message
already present in source can be returned only after restrictive length/content
checks. Dynamic error values, arbitrary stack traces and credential/path-like
content are withheld. Preparation, missing dependencies and RDS I/O remain
hard operational failures instead of being disguised as repairable business logic.

New-protocol dependency failures carry up to eight static references and 32
eligible package names within eight KiB, with explicit omission counts. They
distinguish undeclared/unavailable packages, unexported/private members, forbidden
operations and unresolved functions. For example, an eligible package can still
have a requested function that is not exported. Repairs receive that exact
reference, not advice to install the package or change the accepted formula.
Call arguments and arbitrary namespace-load errors are never included.

Capture an authoring error to inspect its retained diagnostics without entering
the debugger:

```{r custom-authoring-diagnostics, eval=FALSE}
result <- tryCatch(
  create_custom_model(
    instructions = "Use the same calendar month in the prior year.",
    llm = llm, input_data = data_tbl,
    combo_variables = "id", target_variable = "value",
    date_type = "month", forecast_horizon = 3,
    approval = "automatic"
  ),
  error = function(error) error
)
feedback <- custom_model_feedback(result)
print(feedback[c("status", "stage", "code", "next_action", "attempts", "questions")])
if (identical(feedback$status, "approved")) custom <- result
```

Feedback schema 1 covers `approved`, `review_required`, `needs_input`, and
`failed`. It includes available candidate/contract/request identities, resolved
inputs and bounded safe details without dumping source, training rows or raw
exception text. Questions are labelled provider proposals. NULL counts mean
unavailable evidence, not unlimited retries; approved models do not retain
attempt histories. `needs_user_input` is true for questions and manual review.
Preflight failures without a draft/model have unknown approval policy (`NULL`);
known error stages are retained. Errors still raise normally and are never usable models.
Full diagnostic history remains separately available in `result$diagnostics`
when an error is captured. Both reports can contain confidential column names,
provider-visible questions or explicitly chosen artifact paths.

Coding agents should present `needs_input` questions to the user and inspect the
exact pending review before requesting consent. Feedback does not authorize
execution, restart an interrupted action, increase budgets, or silently refresh
a stale request. Use the same `create_custom_model(draft = ..., responses = ...)`
flow below. LLM configuration remains per call; no global provider configuration
or hard provider/execution timeout is introduced.

The bundle retains at most 12 compact attempts and one candidate/accepted-contract
snapshot under a one-MiB JSON limit. Oversized snapshots have an explicit omission
flag and retained hashes; validation examples themselves are never truncated.
With `control$draft_path`, diagnostics share the existing immutable revisions; otherwise
the condition carries them in memory. No new default data persistence or cloud
upload occurs. Treat source, instructions and provider-visible examples as
confidential. The snapshot excludes real training rows, fitted objects and Chat.
Current drafts require their diagnostic history and explicit approval policy.

Source repair cannot change the rule, units, parameters, expected answers or
tolerances, and each candidate still needs exact manual or automatic authorization.
The three-attempt limits remain unchanged. Tests cover last-value, mean/offset,
rolling and weighted means, seasonal lookup/growth, explicitly defined mean YoY
growth, driver ratios and conditional rules, plus quarterly and pooled controls.
These are examples of supported ordinary-R logic, not a closed model menu or a
promise that every arbitrary or ambiguous request succeeds. Live-provider
reliability requires separate evaluation; no sandbox or hard timeout is added.

### Deferred Review

Manual noninteractive calls without a callback return a non-enrollable draft at
the next review gate. `control$draft_path` persists state without changing review
delivery; `control$interact` supplies synchronous trusted-callback review.
Resume with the same public function and one response bound to the pending
request. Inspect the complete review before entering any confirmation.

```{r custom-authoring-deferred, eval=FALSE}
draft <- create_custom_model(
  instructions = "Use the mean of all historical targets independently for each series.",
  llm = llm,
  input_data = data_tbl,
  combo_variables = "id", target_variable = "value",
  date_type = "month", forecast_horizon = 3,
  approval = "manual", validation_examples = examples,
  control = custom_model_control(draft_path = "custom-model-reviews")
)
print(draft)
next_state <- draft
if (inherits(draft, "finnts_custom_model_draft")) {
  request <- draft$pending
  if (identical(request$stage, "clarify")) {
    answer <- lapply(request$questions, function(question) readline(paste0(question, ": ")))
  } else {
    cat(request$review$contract$interpretation, "\n", request$review$warning, "\n")
    print(lapply(request$review$contract$examples, function(example) example[c("name", "expected")]))
    answer <- readline("Create model? [y/N/details]: ")
  }
  next_state <- create_custom_model(
    draft = draft, llm = llm,
    responses = list(request_id = request$request_id, answer = answer)
  )
}
```

One yes can return the approved model without `llm`; `details` prints the exact
candidate and returns the same draft without changing its revision. For an edit,
send `list(action = "edit", instructions = "requested change")` as the answer.
An edit without `llm` waits at stage `contract`; a failed candidate without an LLM
waits at stage `source`. Resume either with `llm` for the next bounded proposal.
Expected examples stay fixed during code repairs, and cumulative limits survive
every pause. Changed source requires a fresh combined review and decision.

An explicit `control$draft_path` retains a separate private context snapshot and versioned
review files across R sessions. Resume using the returned object or its exact
`file.path(draft$store, "current.rds")` path. `saveRDS(draft, ...)` alone does not
copy the referenced data. Without `control$draft_path`, temporary context may expire when
R exits. The caller owns access control and retention: source, instructions,
examples and snapshot data may be confidential. No automatic cleanup or upload
occurs. Drafts contain references, not training rows, callbacks, fitted workflows
or Chat sessions; their print methods omit review contents and private paths.

Stale replies, missing or changed context, altered evidence and changed runtime
metadata fail before old consent can be reused. Repeating an accepted identical
response returns saved current state without repeating completed work. Interrupted
actions or corrupt pointers require restoration or a new draft, not guessed
approval. These checks are not authentication or concurrent-writer transactions.
The manual approved envelope is unchanged; only a validated envelope can be enrolled.

Authoring validation uses the currently loaded FinnTS functions in the current R
session, including `devtools::load_all()`. It starts no child process and installs
no package snapshot. Load the intended development code and revalidate after
changing it; FinnTS does not compare loaded functions against on-disk source.
The current draft format records the combined decision. Unsupported development
formats require a new call and fresh review; no compatibility implementation is
retained for them.
Local authoring prepares at most three deterministic series; explicit global
authoring retains the complete supplied cohort. Both enforce at most 10000 total
historical rows, using a separate temporary root for raw data. Existing prepared
runs are read without writes. Explicit hierarchy authoring samples up to three
raw leaf series and retains every derived node; a prepared hierarchy is read
completely. The 10000-row limit applies to all historical node rows, and larger
hierarchies are rejected rather than truncated. The configured provider receives instructions,
column types and fixed examples during generation, not automatically raw Target/regressor
values or business series labels. The return is passive and serializable; it
contains measured summaries but no fitted state, raw data, callbacks or Chat.

Generated R runs in your current session with access to session/environment values
and your account's filesystem/network permissions. There is no sandbox, crash
isolation or reliable automatic interruption; stuck code may require restarting R.
`control$validation_budget` is an elapsed acceptance budget checked after evaluation
returns, not an interrupting timeout. Ordinary options, RNG, working directory
and library paths are restored on normal return/error, but arbitrary global,
environment, filesystem or native-code side effects cannot be rolled back.
Manual mode requires review before consenting; automatic mode explicitly waives
that review. Sample validation is evidence, not a proof of all future
behavior. No packages are installed and authoring does not register the result or
publish a forecast. Select its alias explicitly through `models_to_run` when
supplying `custom_models = list(business_mean = custom)` to standard preparation.

The original bottom-up profile requires these explicit settings:

| Setting | Required Value |
| --- | --- |
| `forecast_approach` | `"bottoms_up"` |
| `recipes_to_run` | `"R1"` |
| Target scale | Original, with `stationary = FALSE` and `box_cox = FALSE` |
| Target cleaning | `clean_missing_values = FALSE`, `clean_outliers = FALSE` |
| `multistep_horizon` | `FALSE` |
| `run_ensemble_models`, `feature_selection`, `inner_parallel` | `FALSE` |
| `negative_forecast` | `TRUE`, so prescribed output is not clipped |
| Storage | Local or mounted paths with `storage_object = NULL` |
| Formats | CSV/RDS/Parquet data and RDS model objects |
| Execution | Sequential or `parallel_processing = "local_machine"` |

`forecast_time_series()` now exposes `stationary`, retaining its existing `TRUE`
default for ordinary calls. The other unsupported transformations are not
silently disabled. Staged calls validate saved preparation settings and later
training controls; the high-level wrapper rejects incompatible explicit settings
before preparing data. The public `set_run_info()` constructor continues to
advertise CSV/Parquet data output; existing RDS-compatible run settings remain
supported by the internal storage path.

Custom workflows preserve Date and Combo plus explicitly declared predictors.
They do not acquire implicit PCA, feature selection, target transforms or hidden
features. Missing required data is an error. Local/global declarations constrain
the enabled run modes; a definition with no compatible enabled mode fails rather
than enabling a different mode. Custom fits use chronological analysis folds and
fixed parameters, without tuning, source regeneration or package installation.
Custom source or incomplete-prediction failures cannot be silently dropped to
make a mixed run appear successful.

A run manifest pins the effective pool, exact source versions, attestations and
preparation context. Reordered equivalent selections can reuse it; changed
versions, approvals or context require a new run. Custom `Model_Name` values keep
their readable aliases, including underscores. Custom `Model_ID` values use
`custom-<full version hash>--<local|global>--R1`, avoiding ambiguity with existing
average-component separators. Saved workflows, fits, predictions and averages
are checked before completion shortcuts; stale versions and foreign components
raise errors instead of being silently repaired.

The ordinary accuracy and plausibility policy still selects the best eligible
candidate. With `average_models = TRUE`, a custom-only run may select an average
of its eligible custom components, but never an unrequested built-in or learned
ensemble. Enrollment does not guarantee selection. Exact single-rule workflows
must explicitly disable averaging; `bottoms_up` alone does not do that. R2,
transformed targets, learned ensembles, Spark and remote standard execution
remain outside this profile.

### Reviewed Hierarchy Rules

Use `create_custom_model(control = custom_model_control(forecast_approach = "standard_hierarchy"), ...)` or
`"grouped_hierarchy"` to review a new local-per-node definition. Its contract
explicitly permits normal reconciliation to adjust the rule's base predictions
before publishing the bottom level. An old bottom-up approval cannot be silently
upgraded; changing hierarchy semantics requires a new version and human approval.

Forecast enrollment must include both compatible custom rules and eligible FinnTS
models in one shared `models_to_run` pool. A custom-only hierarchy request errors
without adding a model or switching the approach. Individual rule validation
during authoring does not require a competitor and does not enroll a forecast.
Different winners at different nodes, including custom winners everywhere, are
valid outcomes of an eligible mixed pool.

This profile is local, univariate, original-scale R1 with no averaging, cleaning,
transforms, learned ensembles or inner parallelism. Every candidate must cover
every required node and forecast key. Missing coverage cannot be replaced by
`snaive`, zero-filled or silently omitted. Existing hts mathematics and quality
selection are reused. Per-series routing, pooled/global hierarchy and Agent
hierarchy remain later work. See the [hierarchy walkthrough](hierarchical-forecasting.html#reviewed-custom-rules)
for explicit controls and the compatibility evidence matrix.

Initial Agent iteration also accepts approved pools through `set_agent_info()`.
Only explicit `models_to_run` enrolls candidates. The Agent retains the same
ranking and average-model accuracy policy, but proposals cannot expand the pool
or change its original-scale R1 representation. Parent and child identities are
checked before same-run reuse and publication. This initial Agent profile is
limited to local/mounted CSV data and RDS objects, sequential or local-machine
execution, with no inner parallelism. Compatible custom updates require a new
identically enrolled parent and explicit `allow_iterate_forecast = FALSE`. They
pin and replay the exact saved choices with unchanged horizon, regressors, modes
and series membership. Existing update acceptance rules still apply, but rejection
or damage raises an error instead of default recovery. Custom rules are never
retuned; mixed retuning reuses their fixed results. Valid update restarts preserve
completed outputs. A built-in-only parent cannot adopt a custom predecessor.
See the [Agent walkthrough](ai-agent.html#approved-custom-models) for the bounded
update workflow and unsupported recovery cases.

## Future-forecast checks

These checks apply to candidate base forecasts before reconciliation, including averages. Candidates with incomplete or non-finite predictions are ineligible. Backtests and future predictions without a supported trend are rejected when their absolute magnitude exceeds 100 times a positive robust historical scale. That scale uses the 95th percentile of absolute levels and robust level/change variation, not just the last observation. A supported trend supplies a pointwise future bound of `100 * max(robust_scale, abs(projected_reference))`, so sustained growth is not rejected merely for accumulating over a long horizon. The projection comes only from history, never from the candidate being screened. Existing finite-negative handling still follows `negative_forecast`, but missing or infinite predictions are no longer replaced with zero.

Level and trend deviations beyond six robust reference scales create soft concerns. The reference uses the latest available `max(12, 3 * seasonal_period, 2 * forecast_horizon)` periods before the historical cutoff. Without supported drift, its path is seasonal naive when a complete seasonal cycle is available, otherwise a recent median level. The robust scale is the largest of the 95th percentile of absolute levels, level MAD, and one-step-change MAD. In that fallback, level-reference width is the larger of the relevant change MAD and 5% of that scale, multiplied by `6 * sqrt(horizon_step)` for the comparison. Supported drift instead supplies a projected center and a width reflecting residual noise and slope variation, described below. Trend checks compare median changes over matched historical and forecast spans, with a robust scale floor; proportional references perform those comparisons on log changes.

With at least two complete finite historical cycles, a detrended seasonal profile can support additional checks. Seasonal strength must be at least 0.6 and the period at least three. For forecasts covering at least one cycle, negative profile correlation or amplitude outside one-third to three times historical amplitude creates a concern. Supported proportional trends use log-scale historical and forecast profiles, so constant proportional seasonality is not confused with growing absolute amplitude. Additive and fallback paths retain the original seasonal calculations. Unsupported checks remain unassessed. These are guardrails, not calibrated prediction intervals or guarantees of future accuracy.

For a shorter horizon with at least three future points, Finn compares only the upcoming phases. Their historical range must exceed both three times the historical phase-residual MAD and 5% of the full historical seasonal amplitude. Finn removes the robust historical deseasonalized trend from the future path, rather than fitting a free trend through those few points. Negative correlation creates a phase concern only when future variation also exceeds that noise threshold. Full-cycle amplitude concern bounds are not applied to a partial cycle.

One or two seasonal cycles remain usable. One cycle supplies a cautious reference, while repeated cycles support stronger comparisons. Weak seasonality, fewer than three future points, low-variation partial phases, and unavailable checks remain unassessed instead of making the series fail. Flat forecasts are not rejected simply because observations are noisy.

| Check | Trigger | Effect | When Unassessed |
| --- | --- | --- | --- |
| Coverage | Missing, duplicate, or unexpected expected scenario/date keys in backtests or future predictions | Hard rejection | Never for a candidate being selected |
| Finite predictions | Any required `NA`, `NaN`, `Inf`, or `-Inf` prediction | Hard rejection | Never |
| Catastrophic magnitude | Absolute backtest or unsupported-trend future prediction exceeds `100 * robust_scale`; supported future step exceeds `100 * max(robust_scale, abs(projected_reference))` | Hard rejection | When historical scale is zero |
| Accuracy availability | No finite weighted MAPE from usable actuals | Hard rejection | Never |
| Level | Forecast departs from the supported trend reference, or seasonal-naive/recent-median fallback, beyond its horizon-adjusted width | Soft concern | Skipped after a hard failure |
| Zero-history level | A nonzero future forecast when the reference history is entirely zero | Soft `unsupported_level` concern | Zero forecasts pass; this is not a ratio-based hard rejection |
| Trend | Median forecast change differs from historical matched-span changes beyond six robust slope scales, using log changes for a supported proportional reference | Soft concern | Fewer than four matched periods, non-finite reference values, incompatible log-domain forecasts, or a hard failure |
| Seasonal amplitude | Future detrended amplitude is below one-third or above three times the historical profile | Soft concern | Insufficient cycles, strength below 0.6, period below three, short forecast, or zero historical profile amplitude |
| Seasonal phase | Future and historical profiles have negative correlation after the appropriate trend adjustment | Soft concern | Insufficient historical cycles or strength, period below three, undefined correlation, or insufficient points or signal for a partial horizon |
| Seasonal amplitude preference | Amplitude difference exceeds the tolerance learned from historical cycle variation | Tie-break only; no additional concern or rejection | Insufficient historical evidence, zero reference amplitude, or insufficient partial-horizon phase signal |

For these checks the default seasonal period is daily `7`, weekly `52`, monthly `12`, quarterly `4`, and yearly `1`. If multiple seasonal periods are configured, the smallest valid integer greater than one is used. The checks do not independently validate every seasonal pattern a model can learn. The trend scale is the larger of historical matched-span slope MAD and `0.05 * working_scale / matched_span`. The working scale is the original normalized robust scale for additive or fallback paths, and the analogous robust scale of normalized log values for proportional paths.

Each assessed soft check has a normalized excess score. Candidate risk is the maximum of the level, trend, and seasonality scores, not their average. The concern count counts the triggered reasons; seasonal phase and amplitude can contribute two reasons. Unassessed checks do not count as concerns.

The separate in-memory `Seasonal_Fidelity` score measures amplitude distortion beyond historical variation. For the phases being assessed, let `reference_amplitude` be the reference profile range, `future_amplitude` the comparable future range, and `cycle_amplitudes` the ranges of centered historical cycle profiles. The tolerance is `max(3 * MAD(cycle_amplitudes), 0.05 * reference_amplitude)`. Fidelity is the finite nonnegative score `max(0, (abs(future_amplitude - reference_amplitude) - tolerance) / reference_amplitude)`. Insufficient evidence gives `NA`, not an artificial zero. A partial-horizon flat average can have an assessed amplitude preference even when its phase correlation is undefined. This score does not change risk or the concern count, and by itself cannot reject reuse, prevent accuracy-goal stopping, or trigger another fit.

## Supported historical growth

A trend reference requires at least `max(12, 3 * seasonal_period)` finite, regularly spaced prepared-history values in the existing reference window. Missing dates are not bridged, and missing values are not dropped to manufacture regular history. If explicit observation flags supplied to the evaluator identify missing or unobserved fitting values, trend support is declined. Existing prepared artifacts do not establish such provenance automatically; retained imputation is still part of their evidence.

Finn compares at most two robust reference forms, without training any additional forecasting model:

- Additive drift uses normalized target values. Drift is the median of same-phase changes divided by the seasonal period, or adjacent changes for nonseasonal yearly data. Phase intercepts are medians after removing that drift, rather than an extrapolation anchored to one last-point spike.
- Proportional drift applies the same method to `log(target) - log(normalization)`. It is considered only when all historical values are strictly positive and exceed `sqrt(.Machine$double.eps) * robust_scale`. Signed, zero, and relatively near-zero history is not shifted, clipped, or transformed with `log1p` to make this form fit.

With period `p`, two chronological prefixes end at `n - 2q` and `n - q`, where `q = max(1, floor(p / 2))`. Each predicts the next `q` historical values and must contain at least `max(8, 2p)` training values. Drift must be nonzero, have the same sign in both prefixes and the full window, and exceed twice the MAD of its per-step slope estimates, with a numerical-precision floor. Each validation block must improve MAE by at least 20% over the no-drift reference on the same dates. Comparisons occur on the original target scale, normalized by a common factor for numerical stability. Proportional drift displaces an accepted additive reference only if it improves additive MAE by at least 20% in both blocks. Zero-error and negligible comparisons retain the simpler reference. These fixed support thresholds are engineering choices, not statistical confidence statements.

For a supported reference, let `sigma_error` be the larger of its residual MAD and a noise floor, and `sigma_slope` the MAD of its per-step drift estimates. The additive floor is 5% of the existing normalized robust scale; the log-space floor is `log1p(0.05)`. At step `h`, the level-comparison width is:

$$
6\sqrt{h\,\sigma_{\mathrm{error}}^2+h^2\,\sigma_{\mathrm{slope}}^2}.
$$

Overlapping slope estimates are not treated as independent observations to shrink uncertainty. The projected center follows the supported drift and upcoming seasonal phase; it does not widen itself in response to candidate forecasts. Reference estimation is shared across candidates through the existing evaluation cache. Nonrepresentable reference projections fall back to the original checks instead of clipping forecasts or removing the magnitude bound.

For example, a positive series with well-supported 2% monthly growth can have a reference near 126.8 after 12 months and 160.8 after 24 months when its current fitted level is 100 and it has no seasonal effect. Continuing that growth need not produce a level or absolute-slope penalty merely because the horizon is longer. A path that accelerates beyond the supported trajectory can still receive a concern or fail its pointwise magnitude bound. This reference is not substituted for the selected model's actual forecast.

A finite zero or negative forecast under a supported proportional reference receives a `level_deviation` soft concern, while log-only slope and seasonal comparisons are unassessed. The forecast is not changed, and this does not introduce a new hard sign veto. Short, irregular, missing, or unstable histories and all-zero series retain the earlier fallback behavior. New scoring applies to newly evaluated predictions; a complete saved winner is still reused without retrospective reassessment. Unexpected regimes and mixed seasonal mechanisms can remain uncertain, and an eligible singleton with soft concerns can still win under the ordinary best-available policy.

## Accuracy allowance

After hard failures are excluded, let `best_wmape` be the smallest eligible weighted MAPE as a fraction. Candidates within `best_wmape + max(0.005, 0.05 * best_wmape)` form the shortlist. Selection prefers lower forecast risk, then fewer concerns. Within each risk-and-concern tie, smaller seasonal amplitude distortion precedes WMAPE only when every tied candidate has a finite assessed fidelity score. Otherwise that tie retains the existing WMAPE ordering. A stable candidate identifier breaks the final tie. Missing seasonal evidence is not rewarded as zero distortion.

For example, with a best eligible WMAPE of 8.0%, the ceiling is 8.5%. A safer 8.3% candidate can beat a questionable 8.0% candidate, but an 8.8% candidate is outside the allowance. At 20% the ceiling is 21%; at 2% it is 2.5%. The absolute allowance is relatively generous for very accurate series and is an initial engineering default, not a calibrated statistical bound.

For ordinary and iterative forecasts, when every shortlisted eligible candidate has soft concerns, Finn selects the best available one and reports them. When all candidates hard-fail, Finn raises an error. Update reuse and default replacements have a stricter acceptance boundary, described below. No policy can guarantee that a good forecast exists for every input.

## Standard forecasts and saved averages

`forecast_time_series()` finishes through `final_models()`. A manual standard workflow uses the same process when `final_models()` is called:

1. Load prepared original-scale actuals and expected backtest/future dates. Future target values never become historical evidence.
2. Exclude candidates with missing or duplicate required keys, non-finite predictions, catastrophic magnitude, or unavailable finite WMAPE. A model with incomplete backtest coverage cannot win or enter a simple average. Learned-ensemble inputs are screened before fitting their existing ensemble specifications.
3. Form the requested simple averages from eligible existing predictions, up to `max_model_average`. A non-finite component is not concealed by averaging with `na.rm = TRUE`.
4. Evaluate the individual, learned-ensemble, and simple-average candidates on the same dates and actuals. Apply the accuracy allowance, then rank by risk, number of concerns, the supported seasonal-amplitude preference within ties, WMAPE, and stable model identifier.
5. Mark the delivered winner `Best_Model = "Yes"`. If it is an average, save that exact combination in the existing average-model artifact. Otherwise rank just the eligible averages by the same policy and save the winner of that subset with `Best_Model = "No"`. This is not necessarily the lowest-WMAPE average. If no eligible average can be formed, there is no required average artifact.
6. For hierarchical forecasts, pass the selected models at every prepared hierarchy node to the existing `hts` solver and publish its bottom-level result. Do not run another future-quality selection after reconciliation. Format intervals and any weekly-to-daily allocation using the existing output workflow.

Individual outputs remain available with their best-model flags, but only the selected simple average is retained, not every computed combination. Standard forecasting selects among models already run; a quality failure does not fit an additional model automatically.

On retry, Finn validates the combined saved outputs before reusing a winner. A `Best_Model` column, an average filename, or a finite completion log alone is insufficient. All individual models marked `"No"` can be correct when a complete saved average is marked `"Yes"`. Missing, incomplete, or ambiguous selection flags cause averages and selection to be rebuilt from existing predictions, without training again. Saved weekly outputs are restored to native cadence before recomputation. Complete winners are not re-evaluated for future quality, and a multi-series retry retains both previously completed and newly finalized results. A saved average whose original component outputs are unavailable requires restoration of those artifacts; it is not silently reconstructed from a different subset.

## Reconciliation

Quality-aware selection happens before reconciliation. Different hierarchy nodes can choose different models or averages. Finn reconciles that selected mixture with the existing `hts::combinef()` nodes/groups, residual weights, and nonnegative settings, then publishes the resulting bottom-level forecasts.

There is no runtime plausibility scoring of reconciled future paths, no switch of the whole output to a uniform reconciled model family, and no replacement of individual bottom rows after the solve. Standard runs may still save per-model reconciled outputs for inspection, but those outputs are not automatically promoted by a post-reconciliation quality ranker. An outer Agent reconciliation solves its selected mixture directly.

Reconciled backtests provide normal bottom-level WMAPE reporting and run comparison. The solver consumes selected forecast values, hierarchy structure, and residual-based weights, not quality rankings. A resumed run does not require every source node's prior ranking or repeat its future-quality assessment. Actual complete selected source forecasts are still required when a solve must run. Weekly source forecasts are evaluated at weekly cadence during selection, and a non-finite value on any daily-expanded forecast row remains non-finite when its weekly key is restored.

Reconciliation can dampen, spread, or amplify a problem. Better base forecasts can improve bottom-level output, but passing the source checks is not a guarantee that every reconciled path is plausible. Solver and ordinary artifact errors remain errors. Test-only audits can measure the resulting forecasts without changing production selection.

## Iterative Agent forecasts

`iterate_forecast()` runs `final_models()` within each submitted forecast run. Iteration ranking starts with the earliest minimum-WMAPE result in the current Agent version. Later eligible iterations within 10% relative of that WMAPE can become the preferred search context when their `model_avg_wmape` is strictly lower; the lowest such average wins, with earlier ties retained. Missing or non-finite averages do not create an improvement. This is separate from the within-iteration 0.5-percentage-point/5% allowance: risk, concern count, and seasonal fidelity are not reassessed to rank past iterations.

For local iterations, `model_avg_wmape`, `model_median_wmape`, and `model_std_wmape` summarize the individual model candidates' backtest WMAPEs, excluding simple-average forecasts. For example, ARIMA may remain best after adding an external regressor while the multivariate models improve. A lower average preserves that promising search direction because further iterations may make those models the winners; it is not a guarantee of future accuracy. Global summary fields retain their original meanings: overall run WMAPE for mean and median, and zero spread. Statistics use already-loaded current backtests; history comparisons reuse recorded metrics without reloading past forecasts. Unavailable local statistics are not fabricated from the selected model's score.

Choosing an iteration's settings for further optimization is distinct from replacing a saved forecast: a better saved local forecast remains protected. Global promotion is decided once for the complete iteration, then all saved global winners move together even if one series individually worsens. All globally selected series must reference one `best_run_name`; they may still use different models or averages from that iteration. A partial global evaluation cannot promote only its successful series. Reload, finalization, and update reject mixed global iteration metadata, including inconsistent state left by interrupted writes, rather than silently using multiple global iterations. These per-series writes are not a storage transaction. Agent comparisons, saved WMAPE, and accuracy-goal decisions retain four-decimal precision.

Normal accuracy-goal stopping and local optimization routing use completeness and WMAPE, without another soft-quality veto. Consequently, an iteration winner with soft concerns can beat an earlier winner or meet the accuracy goal; hard-invalid and incomplete results still cannot claim successful completion. Rejected runs consume iteration budget. At the limit, an eligible best-available result may be retained with concerns; if none exists, Finn fails or uses an already enabled local phase for unresolved global series. Final outer reconciliation does not initiate another quality-selection or refitting loop. Avoided evaluator calls and artifact reads are covered by tests; no fixed runtime reduction is promised.

For ordinary Agent iterations, the goal and saved-forecast comparisons use selected
completed-output backtests, including daily-expanded rows when weekly forecasts
are delivered daily. Within-run candidate selection stays at native cadence.
Those scores are not interchangeable: the zero-actual convention and rounding can
produce different WMAPEs after daily allocation. Future target placeholders never
contribute to either accuracy calculation.

## Updated Agent forecasts

`update_forecast()` follows a different fitting path but uses the same evaluator:

1. Recover the previously selected single model or average for each series from its saved source forecasts. All globally selected series must reference one winning global iteration; mixed iteration metadata fails before any global refit. Refit only the union of required components from that run with their saved settings; do not average all models listed in the earlier request. Different series or hierarchy source nodes can retain different subsets within that one iteration. Missing or ambiguous selected identities or saved fits require restoring the original artifacts. Reused fits do not call `final_models()` to search over a new candidate pool.
2. Assemble component and reused-average forecasts without reconciling them. Check every required component for hard eligibility and the selected reused combination for applicable soft concerns. Repeat before reconciliation after any existing WMAPE-triggered retuning. Passing reuse keeps that original chosen combination; it does not silently swap to another component.
3. Deduplicate quality-rejected current series into the existing new-series/default-local forecast path. Healthy siblings are retained and removed series are excluded. Quality-only rejections do not consume the ordinary execution-failure cap of `max(10, ceiling(0.20 * existing_series))`.
4. Run `final_models()` for each default replacement, then require the replacement to pass all applicable quality checks. Short-history checks that cannot be assessed do not count as failures. Good backtest accuracy alone cannot publish a still-questionable replacement.
5. An inner reused hierarchical candidate needs its complete source-node forecast set. If a node is rejected or the hierarchy is incomplete, do not solve a partial hierarchy; send the candidate's covered current series through the existing default-local path. Accepted hierarchies reconcile only the selected rows, not both components and their average together. Existing-format source forecast files retain selected component identities for subsequent updates, not a required diagnostic ranking snapshot.
6. If outer reconciliation is required, reconcile the selected forecasts once and publish its result without a post-quality switch or late default reforecast. A default replacement rejected before reconciliation raises an explicit quality error and cannot trigger repeated fitting. Stable default run identities and an acceptance/rejection field in existing run logs preserve that boundary across restarts.

The optional accuracy-degradation `allow_iterate_forecast` workflow remains separate. Checks after refit or retune assess newly generated predictions, not old iteration winners. A default's saved acceptance or rejection prevents repeating that acceptance decision after it has completed; an interrupted acceptance step can still assess the new output once. Quality recovery itself does not ask the LLM to waive a check. Default refitting can increase compute or configured-provider cost, and it is a bounded recovery attempt rather than a promise of a satisfactory forecast.

Update accuracy keeps the existing native-cadence aggregate used for refit/retune
and aggregate logging. Completed-output backtests supply per-series saved-winner
accuracy and local model-pool statistics. This distinction preserves the update
workflow's metric meanings when weekly output is expanded to daily rows.

## Evidence and limitations

Developer tests compare two paths on the same fake candidate outputs: accuracy-only selection followed by real `hts`, and quality-aware selection/averaging followed by real `hts`. Held-out future truth is used only by the tests, never supplied to the selector. They measure bottom-level error and shape, including effects on uninjected siblings. Small regressions run with package tests; broader stress and full-catalogue averaging timings are separate developer experiments.

A separate bounded, in-memory corner-test matrix checks exact prediction keys, signed/zero/missing-actual accuracy, threshold boundaries, seasonal alignment, optional-score ties, weekly group isolation, and replacement-selector contracts. Fixed-seed cases also check independent accuracy and ranking expectations, row-order invariance, and the inability of a hard-invalid candidate to displace a valid winner. These tests do not fit forecast models or run additional reconciliation, and passing them is not proof that every possible input or future regime change is handled correctly.

Earlier experiments exposed half-amplitude averages and short-horizon phase reversals that the original concern checks did not distinguish. Focused regressions now prefer an intact available seasonal candidate within the accuracy allowance and assess informative partial horizons using repeated historical evidence. Controls retain naturally varying seasonal amplitudes within the history-adaptive tolerance. These results do not calibrate a guarantee: one-cycle histories, fewer than three future points, or low-signal partial phases can still retain reversed timing because assessment is unsupported. A plausible alternative outside the WMAPE allowance cannot win a soft-risk or fidelity comparison. Unexpected regime changes are not known in advance, and the existing nonnegative solver floor can produce small positive values for all-zero histories. None of these limits invokes a post-reconciliation replacement model.

The number of simple averages grows as `choose(N, 2) + choose(N, 3)` when `max_model_average = 3`. The normal smoke test uses four mixed model/recipe candidates and independently checks all ten pair/triple averages. The separate developer timing workload uses every supported model/recipe output and reports averaging and evaluation separately, with fresh artifacts for every trial. Earlier timing and broad stress measurements predate the refined seasonal ranking; they are not new measurements of this revision. No new elapsed-time assertion or production timeout is added. Runtime depends on candidate count, horizon, backtests, storage, and the machine; measurements should precede any future regression limit.

## Historical actuals

Checks use the existing prepared data for each series, preferring R1 when prepared and otherwise normalizing R2's `Horizon == 1` rows. Differencing and Box-Cox are reversed using existing metadata. `Target_Original` is used when available so outlier cleaning does not change the actuals used for evaluation; otherwise `Target` is used. Prior missing-value imputation is retained. Recipe and metadata files are read by exact path, without directory listing, and history is reused in memory across candidates.

Selection requires at least one candidate and at least one finite actual inside the recent reference window. An empty candidate pool or a reference window containing only missing/non-finite actuals raises a clear input error. Finite observations farther back, or future target values, cannot silently supply that window's evidence. Partly missing windows remain usable when a finite value, including zero, is present; unavailable seasonal checks remain unassessed, and missing backtest actuals continue to receive no accuracy weight.

Original and cleaned targets have separate inverse-differencing starting values. Older second-order original-target artifacts missing those values must be regenerated from original input; already saved forecast outputs remain readable.

Rankings and explanations are held in memory. No additional diagnostic or reference files are written. The selector has an internal replaceable function contract; a public per-run custom-evaluator API is not included in this release.

## Backtest weighted MAPE

A pointwise absolute percentage error is produced first, rounded to four decimal places, then weighted by the absolute actual value across expected backtest rows. The existing convention replaces zero actuals with `0.1` for this calculation. Original actuals are matched by date; missing/non-finite actuals do not contribute weight, and a candidate without a finite accuracy score is ineligible. Overlapping backtests keep their separate scenario/horizon observations. The positive-target example below illustrates the weighting.

```{r, echo = FALSE, message = TRUE}
suppressMessages(library(dplyr))

message("Simple Back Test Results")
back_test_tbl <- tibble(
  Combo = c("Country_1", "Country_1", "Country_1", "Country_1", "Country_1", "Country_1", "Country_1", "Country_1", "Country_1", "Country_1"),
  Date = c("2020-01-01", "2020-02-01", "2020-03-01", "2020-04-01", "2020-05-01", "2020-01-01", "2020-02-01", "2020-03-01", "2020-04-01", "2020-05-01"),
  Model = c("arima", "arima", "arima", "arima", "arima", "ets", "ets", "ets", "ets", "ets"),
  FCST = c(9, 23, 35, 41, 48, 7, 22, 29, 42, 53),
  Target = c(10, 20, 30, 40, 50, 10, 20, 30, 40, 50)
) %>%
  dplyr::mutate(
    MAPE = round(abs(Target - FCST) / abs(Target), 4),
    Date = as.Date(Date)
  ) %>%
  dplyr::group_by(Combo, Model) %>%
  dplyr::mutate(
    Target_Total = sum(abs(Target)),
    Percent_Total = abs(Target) / Target_Total
  ) %>%
  dplyr::ungroup()

print(back_test_tbl)

message("")
message("Overall Model Accuracy by Combo")

suppressMessages(best_model <- back_test_tbl %>%
  dplyr::group_by(Combo, Model) %>%
  dplyr::mutate(Weighted_MAPE = MAPE * Percent_Total) %>%
  dplyr::summarise(
    MAPE = mean(MAPE),
    Weighted_MAPE = sum(Weighted_MAPE)
  ) %>%
  dplyr::ungroup())

print(best_model)
```

During the simple back test process above, arima seems to be the better model from a pure MAPE perspective, but ETS ends up being the winner when using weighted MAPE. The benefits of weighted MAPE allow finnts to find the optimal model that performs the best on the biggest components of a forecast, which comes with the added benefit of putting more weight on more recent observations since those are more likely to have larger target values then ones further into the past. Another way of putting more weight on more recent observations is how Finn overlaps its back testing scenarios. This means the most recent observations are tested for accuracy in different forecast horizons (H=1, H=2, etc). More info on this in the back testing vignette.

Users can evaluate the retained model outputs with their own metrics and choose another model. The internal selector is replaceable, but this release does not expose a public per-run evaluator or persist custom selection functions.
