---
title: "Hierarchical Forecasting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hierarchical-forecasting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The finnts package leverages the great work of the [hts](https://github.com/earowang/hts) package. It's currently retired but we think old dogs can still learn new tricks! It's still of great use to finnts and allows for both standard and grouped hierarchical forecasting.

### Standard Hierarchy

A standard hierarchy of data is pretty straight forward. Each combo variable can be aggregated into the next combo variable when producing forecasts at higher aggregations of the data. This structure resembles a pyramid, with the bottom being the lowest granularity of time series and the top being a single time series of the grand total of the data. Below is a good example with geographical combo variables that can be aggregated into one another when building a standard hierarchy forecast.

```{r, echo = FALSE, message = TRUE}
suppressMessages(library(dplyr))

message("Standard Hierarchical Time Series Data")

hts <- tibble(
  Continent = c("North America", "North America", "North America", "North America", "North America", "North America", "North America", "North America", "North America"),
  Country = c("United States", "United States", "United States", "United States", "United States", "United States", "Mexico", "Mexico", "Mexico"),
  City = c("Kansas City", "Kansas City", "Kansas City", "Seattle", "Seattle", "Seattle", "Mexico City", "Mexico City", "Mexico City"),
  Date = c("2020-01-01", "2020-02-01", "2020-03-01", "2020-01-01", "2020-02-01", "2020-03-01", "2020-01-01", "2020-02-01", "2020-03-01"),
  Target = c(100, 250, 320, 80, 200, 270, 50, 80, 120)
) %>%
  dplyr::mutate(Date = as.Date(Date))

print(hts)
```

In the above example, "City" was the lowest level of the hierarchy, which feeds into "Country", which then feeds into "Continent". For a non-agentic run, Finn forecasts each model at the prepared hierarchy levels, selects the best forecast at each level, and reconciles both the selected combination and the individual models back down to the lowest level. `get_forecast_data(run_info)` returns every successfully saved per-model reconciled forecast plus `Best-Model`. Models that fail to produce a reconciled artifact are not fabricated in the result. Filter `Best_Model == "Yes"` to retain only the reconciled selected forecast.

Agentic runs use a different output contract. During `iterate_forecast()`, different series can run and select different model sets, recipes, or averages. Finn therefore reconciles the selected forecasts together, and `get_agent_forecast(agent_info)` returns only that final reconciled `Best-Model` forecast, not a reconciled comparison table for every model. `Best-Model` can combine different winners across series; it does not mean that the same model family won everywhere. This distinction applies to both standard and grouped hierarchies and to Agent forecast updates.

### Grouped Hierarchy

Grouped hierarchies are very different than the traditional hierarchy approach described above. There are some data sets that can be aggregated in various ways, meaning they need to follow another approach the hts package calls "grouped". A good example is a data set that contains historical time series by geography, customer segment, and product.

```{r, echo = FALSE, message = TRUE}
suppressMessages(library(dplyr))

message("Grouped Hierarchical Time Series Data")

gts <- tibble(
  Country = c("United States", "United States", "United States", "United States", "United States", "United States", "Mexico", "Mexico", "Mexico", "Mexico", "Mexico", "Mexico"),
  Segment = c("Enterprise", "Enterprise", "Enterprise", "Public Sector", "Public Sector", "Public Sector", "Enterprise", "Enterprise", "Enterprise", "Enterprise", "Enterprise", "Enterprise"),
  Product = c("Coffee", "Coffee", "Coffee", "Coffee", "Coffee", "Coffee", "Coffee", "Coffee", "Coffee", "Tea", "Tea", "Tea"),
  Date = c("2020-01-01", "2020-02-01", "2020-03-01", "2020-01-01", "2020-02-01", "2020-03-01", "2020-01-01", "2020-02-01", "2020-03-01", "2020-01-01", "2020-02-01", "2020-03-01"),
  Target = c(10, 20, 30, 5, 8, 11, 20, 23, 27, 50, 55, 60)
) %>%
  dplyr::mutate(Date = as.Date(Date))

print(gts)
```

It would be hard to aggregate the above data in a traditional hierarchy. The same products are found in different segments and countries, also the same segments are found in multiple countries. Finn will follow a similar modeling process as the one described for a traditional hierarchy, but instead will create forecasts at the below levels.

-   Grand Total: model "Target" across the sum of "Country", "Segment", and "Product"
-   Country: model "Target" across the sum of "Country". Creating forecasts for the grand total of "United States" and "Mexico"
-   Segment: model "Target" across the sum of "Segment". Creating forecasts for the grand total of "Enterprise" and "Public Sector"
-   Product: model "Target" across the sum of "Product". Creating forecasts for the grand total of "Coffee" and "Tea".

### Reviewed Custom Rules

Standalone hierarchy runs can apply a reviewed custom rule independently to every
prepared node, including aggregates, then use FinnTS's regular hts reconciliation
to publish the lowest level. Review must explicitly permit this adjustment:
reconciled values are not guaranteed to equal the rule's original base predictions.

An HTS forecast using custom models requires a shared, explicitly enrolled mixed
custom/FinnTS candidate pool. Custom-only requests must use `bottoms_up` with a
compatible bottom-up definition. Finn does not silently add competitors or switch
approaches. Eligibility depends on requested candidates, not winners; custom
models may win everywhere. Existing built-in-only hierarchy behavior is unchanged.

```{r reviewed-custom-hierarchy, eval=FALSE}
custom <- create_custom_model(
  instructions = paste(
    "For each hierarchy node, forecast the mean of all historical targets.",
    "Allow normal FinnTS reconciliation to adjust base forecasts."
  ),
  llm = llm, input_data = data_tbl,
  combo_variables = c("Region", "Product"), target_variable = "value",
  date_type = "month", forecast_horizon = 3,
  model_type = "local", name = "business_mean",
  forecast_approach = "standard_hierarchy"
)

run_info <- set_run_info(data_output = "csv", object_output = "rds")
forecast_time_series(
  run_info, data_tbl, c("Region", "Product"), "value", "month", 3,
  forecast_approach = "standard_hierarchy",
  custom_models = list(business_mean = custom),
  models_to_run = c("business_mean", "meanf"),
  recipes_to_run = "R1", stationary = FALSE,
  clean_missing_values = FALSE, clean_outliers = FALSE,
  run_local_models = TRUE, run_global_models = FALSE,
  run_ensemble_models = FALSE, average_models = FALSE,
  feature_selection = FALSE, inner_parallel = FALSE,
  negative_forecast = TRUE, weekly_to_daily = FALSE
)
```

The input must contain a complete finite leaf panel over the requested historical
window. This custom profile does not silently zero-fill missing observations.
Raw authoring samples up to three deterministic leaf series and keeps all their
derived nodes; prepared authoring reads the whole existing hierarchy without
modifying it. Both enforce a 10000 historical-node-row bound and validate every
captured node. Deferred review uses the same gates and persists exact source
references. Individual authoring validation does not enroll a custom-only HTS run.

Old bottom-up definitions and hashes remain valid but do not acquire hierarchy
permission. New hierarchy definitions bind one approach, per-node application
and permission for reconciliation. The approved envelope and registry format are
unchanged. Run manifests additionally bind topology and input artifacts. Changed
context or missing base candidates cannot be treated as a completed restart.

Custom hierarchy requires `average_models = FALSE`, no external regressors, no
target transformations or cleaning, and no pooled/global fitting. Weekly output
stays weekly. Every enrolled candidate must have complete finite node/fold/date
coverage before reconciliation; missing nodes are not filled from `snaive` and
failed candidates are not silently skipped. Normal weights and selection policy
remain unchanged. Per-model results retain their readable aliases and full
version IDs; the selected mixture is published as `Best-Model`.

The following matrix distinguishes implementation from qualification evidence.
It does not imply every cross-product of formats, cadences and workers was tested.

| Capability | Current Evidence or Boundary |
| --- | --- |
| Standard/grouped, local R1, monthly CSV | Implemented-tested: real mixed forecasts, strict coverage and exact-version restarts |
| Grouped monthly CSV, two local-machine workers | Implemented-tested: sequential parity, two worker PIDs and exact installed namespace |
| Year/quarter/month/week/day at native cadence | Implemented-tested: independent reconciliation projection and date-key checks; complete nonmonthly pipeline qualification remains open |
| RDS and Parquet data, RDS objects | Implemented-tested: preparation, context capture and exact-path reload; full forecasting/worker combinations remain unverified |
| Raw/prepared deferred hierarchy review | Implemented-tested: real installed TEST/APPROVE children, complete-node evidence and authored mixed forecast |
| Other installed compatible local R1 FinnTS competitors | Inert workflow/dependency checks implemented; individual model-family qualification remains open |
| Custom-only HTS | Disallowed by explainability policy; use compatible bottom-up definitions |
| Foundation-model competitors | Not qualified for this custom hierarchy profile; rejected rather than counted as token competitors |
| Drivers, pooling, R2, transformed targets, hierarchy averages, weekly expansion | Remaining work; rejected by the current profile |
| Agent hierarchy and hierarchical new-data updates | Remaining work; current custom Agent behavior remains bottom-up only |
| Remote/Spark, other platforms, minimum dependency versions and live providers | Not qualified here; separate approval and validation required |

The per-series routing interface is deferred. Future user-specific model
restrictions will require `bottoms_up`; ordinary node-level winner differences
and internal worker partitions are not routing restrictions. Users requiring
exact single-rule bottom-up output must also explicitly disable averaging.
Trusted source and RDS are not sandboxed or authenticated approval. Offline
provider fixtures verify control flow, not live generation quality.

### External Regressors

The aggregation process is automatically calculated for each external regressor, depending on how the regressor maps to the combo variables. If a regressor is unique for each time series, then the standard aggregation process is implemented (same as the process for the target variable). If the regressor repeats the same values across all time series, then only one global copy is retained and applied to all new aggregated time series. If the regressor specifically maps to one or more combo variables, then the relationship is respected while still being able to aggregate to the total level. Aggregations of drivers are always summed. 

The only limitation is when an external regressor maps to the middle of a standard hierarchy (not at the global level or bottom level), in this case the regressor will be summed at the global level across the hierarchy. This is due to limitations of aggregation naming in the hts package used in finnts.

Explore the final results of the aggregations by seeing the end result using `get_prepped_data()`.

### Spark Parallel Processing

There is also a small limitation when doing hierarchical forecasting using spark as the parallel computing back-end. The hts package finnts uses cannot handle spark data frames. So behind the scenes finnts has to bring all of the data into memory on the spark cluster driver node in order to create the hierarchies. Please keep that in mind. If you have a lot of historical data that is greater than the RAM on the spark driver node, please consider using a different VM size with more memory. The same issue holds true when reconciling the hierarchical forecast, where all of the forecasts across all time series for a specific model is loaded into the same executor node to be reconciled. Because of these limitations, we are now exploring other options outside of the hts package, including the building of our own hierarchical forecasting package. Stay tuned.

### Corner Cases in Forecast Reconciliation

Some forecasts can be widely off from the target value. This can cause issues when finnts tries to use historical back testing residuals as the initial weights in hts to reconcile the forecast at various aggregations of your data down to the lowest level. To prevent reconciliation issues, finnts will shrink these residuals to more appropriate numbers that still convey they are considerably off from the initial target values, but not astronomical. For example a residual value of 1,000,000,000,000 compared to a target value of 100 might get shrunk down to 900, since keeping the initial residual will throw the weighted reconciliation process out of whack and cause issues. Most forecasts in finnts will not have this problem, but if your data is very noisy then this fix will help prevent any issues when trying to run hierarchical forecasts in finnts. 
