---
title: "Hierarchical Forecasting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hierarchical-forecasting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The finnts package leverages the great work of the [hts](https://github.com/earowang/hts) package. It's currently retired but we think old dogs can still learn new tricks! It's still of great use to finnts and allows for both standard and grouped hierarchical forecasting.

### Selecting a Hierarchy

Use `detect_hierarchy()` before a non-agentic forecast to choose a structurally
valid approach. A unique bottom-level ID alone does not prove nesting: intermediate
labels can still occur under multiple parents. Detection checks every adjacent
parent-child relationship in the same cardinality-based order used by the engine.

```{r, eval = FALSE}
decision <- finnts::detect_hierarchy(
  input_data = data,
  combo_variables = c("Region", "Country", "Site"),
  target_variable = "Revenue",
  combo_cleanup_date = as.Date("2025-08-01"),
  hist_end_date = as.Date("2026-08-01")
)
decision$forecast_approach
decision$conflicts
# Pass decision$forecast_approach as forecast_approach to prep_data(),
# using the same input data and cleanup settings.
```

The function returns diagnostics in memory and does not save files or change
the caller's input. Without a cleanup date it considers all supplied tuples;
with a cleanup date it uses the existing target-sum cleanup policy and excludes
future targets from activity. One retained series selects `bottoms_up`, and no
retained series produces an error. Character-like labels are trimmed consistently
with preprocessing; invalid labels and ambiguous combo identities fail explicitly.
Collect Spark data frames before detection.

Detection describes observed data, not unobserved business relationships.
Inspect `hierarchy_order` and `conflicts` when validating the result. Existing
explicit `prep_data()` settings are not overridden. Agent setup and EDA share
the structural analysis, with an intentional single-column bottoms-up policy.
Adopting
a different approach changes the prepared aggregates and hierarchy metadata;
use a new run rather than reusing artifacts from a different hierarchy.

New Agent versions with only one retained series use `bottoms_up`, even with
multiple combo columns. EDA reports `none`, and global reasoning allows only
bottoms-up for this scope. Pre-expanded hierarchy-level `ID` inputs also use
bottoms-up internally, but retain their configured outer reconciliation.

Resuming an existing Agent version preserves its saved forecast approach and
previously uploaded inputs; a package upgrade does not reclassify that version.
API request-ID retries select the version belonging to that request, even if a
newer request has since created another version.
Logged settings such as dates, horizon, regressors, and forecast flags must still
match. Supplied replacement data values are not uploaded on a resume; create a
new version to use changed history, drivers, or series. Missing required saved
artifacts remain errors rather than triggering a hierarchy migration.
Legacy bottoms-up run logs without `allow_hierarchical_forecast` cannot reveal
whether expansion was disabled or the single-column policy selected bottoms-up;
these retain the existing flag-comparison fallback. A saved hierarchical
approach proves expansion was enabled. No new log columns are added.

A new version uses current detection on its cleaned population. If its outer
approach differs from the latest completed predecessor (for example, standard
to grouped, or hierarchical to single-series bottoms-up), `update_forecast()`
rejects reuse and directs the user to `iterate_forecast()`. The
`allow_iterate_forecast` performance fallback does not bypass this compatibility
check. Existing predecessor artifacts are not rewritten.

### Standard Hierarchy

Hierarchical external regressors retain the existing additive convention.
A level-mapped driver must have at most one nonmissing value for each source
tuple and date. Inference first checks single columns, then progressively wider
tuples only if no narrower grain is valid. At the first valid width, it selects
the grain with the fewest distinct grain/date/value tuples; equally coarse
candidates must identify equivalent source partitions.
Inference checks at most 1,024 candidate grains per driver. Each candidate-width
batch is evaluated completely or rejected before scanning. If the search budget
is exhausted, preparation fails before hierarchy artifacts are written; it does
not silently fall back to bottom-level summation. Reduce redundant combo columns
or prepare the driver at its known source grain before retrying.
Aggregates count each source tuple once within the node's actual
bottom-series membership, including separate tuples with equal numeric values.
Original bottom observations are preserved; missing-only level aggregates stay
missing. Date-global drivers are copied, while full-bottom drivers retain the
existing bottom-level summation and missing-value handling. Equally coarse
candidate levels must describe the same partition or preparation fails with an
ambiguity error. Observed equality cannot establish business meaning: validate
the source grain and additive assumption before preparing a production forecast.
These corrections can change newly prepared driver values. Use a new run or
Agent version; existing saved inputs and fitted models are not migrated.

A standard hierarchy of data is pretty straight forward. Each combo variable can be aggregated into the next combo variable when producing forecasts at higher aggregations of the data. This structure resembles a pyramid, with the bottom being the lowest granularity of time series and the top being a single time series of the grand total of the data. Below is a good example with geographical combo variables that can be aggregated into one another when building a standard hierarchy forecast.

```{r, echo = FALSE, message = TRUE}
suppressMessages(library(dplyr))

message("Standard Hierarchical Time Series Data")

hts <- tibble(
  Continent = c("North America", "North America", "North America", "North America", "North America", "North America", "North America", "North America", "North America"),
  Country = c("United States", "United States", "United States", "United States", "United States", "United States", "Mexico", "Mexico", "Mexico"),
  City = c("Kansas City", "Kansas City", "Kansas City", "Seattle", "Seattle", "Seattle", "Mexico City", "Mexico City", "Mexico City"),
  Date = c("2020-01-01", "2020-02-01", "2020-03-01", "2020-01-01", "2020-02-01", "2020-03-01", "2020-01-01", "2020-02-01", "2020-03-01"),
  Target = c(100, 250, 320, 80, 200, 270, 50, 80, 120)
) %>%
  dplyr::mutate(Date = as.Date(Date))

print(hts)
```

In the above example, "City" was the lowest level of the hierarchy, which feeds into "Country", which then feeds into "Continent". For a non-agentic run, Finn forecasts each model at the prepared hierarchy levels, selects the best forecast at each level, and reconciles both the selected combination and the individual models back down to the lowest level. `get_forecast_data(run_info)` returns every successfully saved per-model reconciled forecast plus `Best-Model`. Models that fail to produce a reconciled artifact are not fabricated in the result. Filter `Best_Model == "Yes"` to retain only the reconciled selected forecast.

Agentic runs use a different output contract. During `iterate_forecast()`, different series can run and select different model sets, recipes, or averages. Finn therefore reconciles the selected forecasts together, and `get_agent_forecast(agent_info)` returns only that final reconciled `Best-Model` forecast, not a reconciled comparison table for every model. `Best-Model` can combine different winners across series; it does not mean that the same model family won everywhere. This distinction applies to both standard and grouped hierarchies and to Agent forecast updates.

### Grouped Hierarchy

Grouped hierarchies are very different than the traditional hierarchy approach described above. There are some data sets that can be aggregated in various ways, meaning they need to follow another approach the hts package calls "grouped". A good example is a data set that contains historical time series by geography, customer segment, and product.

```{r, echo = FALSE, message = TRUE}
suppressMessages(library(dplyr))

message("Grouped Hierarchical Time Series Data")

gts <- tibble(
  Country = c("United States", "United States", "United States", "United States", "United States", "United States", "Mexico", "Mexico", "Mexico", "Mexico", "Mexico", "Mexico"),
  Segment = c("Enterprise", "Enterprise", "Enterprise", "Public Sector", "Public Sector", "Public Sector", "Enterprise", "Enterprise", "Enterprise", "Enterprise", "Enterprise", "Enterprise"),
  Product = c("Coffee", "Coffee", "Coffee", "Coffee", "Coffee", "Coffee", "Coffee", "Coffee", "Coffee", "Tea", "Tea", "Tea"),
  Date = c("2020-01-01", "2020-02-01", "2020-03-01", "2020-01-01", "2020-02-01", "2020-03-01", "2020-01-01", "2020-02-01", "2020-03-01", "2020-01-01", "2020-02-01", "2020-03-01"),
  Target = c(10, 20, 30, 5, 8, 11, 20, 23, 27, 50, 55, 60)
) %>%
  dplyr::mutate(Date = as.Date(Date))

print(gts)
```

It would be hard to aggregate the above data in a traditional hierarchy. The same products are found in different segments and countries, also the same segments are found in multiple countries. Finn will follow a similar modeling process as the one described for a traditional hierarchy, but instead will create forecasts at the below levels.

-   Grand Total: model "Target" across the sum of "Country", "Segment", and "Product"
-   Country: model "Target" across the sum of "Country". Creating forecasts for the grand total of "United States" and "Mexico"
-   Segment: model "Target" across the sum of "Segment". Creating forecasts for the grand total of "Enterprise" and "Public Sector"
-   Product: model "Target" across the sum of "Product". Creating forecasts for the grand total of "Coffee" and "Tea".

### External Regressors

The aggregation process is automatically calculated for each external regressor, depending on how the regressor maps to the combo variables. If a regressor is unique for each time series, then the standard aggregation process is implemented (same as the process for the target variable). If the regressor repeats the same values across all time series, then only one global copy is retained and applied to all new aggregated time series. If the regressor specifically maps to one or more combo variables, then the relationship is respected while still being able to aggregate to the total level. Aggregations of drivers are always summed. 

Intermediate standard-hierarchy nodes use their actual bottom-series membership,
even when `hts` assigns synthetic node names. Each node sums only its own source
entities, rather than inheriting a grand-total driver value.

Explore the final results of the aggregations by seeing the end result using `get_prepped_data()`.

### Spark Parallel Processing

There is also a small limitation when doing hierarchical forecasting using spark as the parallel computing back-end. The hts package finnts uses cannot handle spark data frames. So behind the scenes finnts has to bring all of the data into memory on the spark cluster driver node in order to create the hierarchies. Please keep that in mind. If you have a lot of historical data that is greater than the RAM on the spark driver node, please consider using a different VM size with more memory. The same issue holds true when reconciling the hierarchical forecast, where all of the forecasts across all time series for a specific model is loaded into the same executor node to be reconciled. Because of these limitations, we are now exploring other options outside of the hts package, including the building of our own hierarchical forecasting package. Stay tuned.

### Corner Cases in Forecast Reconciliation

Some forecasts can be widely off from the target value. This can cause issues when finnts tries to use historical back testing residuals as the initial weights in hts to reconcile the forecast at various aggregations of your data down to the lowest level. To prevent reconciliation issues, finnts will shrink these residuals to more appropriate numbers that still convey they are considerably off from the initial target values, but not astronomical. For example a residual value of 1,000,000,000,000 compared to a target value of 100 might get shrunk down to 900, since keeping the initial residual will throw the weighted reconciliation process out of whack and cause issues. Most forecasts in finnts will not have this problem, but if your data is very noisy then this fix will help prevent any issues when trying to run hierarchical forecasts in finnts. 
