---
title: "Hierarchical Forecasting"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hierarchical-forecasting}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The finnts package leverages the great work of the [hts](https://pkg.earo.me/hts/) package. It's currently retired but we think old dogs can still learn new tricks! It's still of great use to finnts and allows for both standard and grouped hierarchical forecasting.

### Standard Hierarchy

A standard hierarchy of data is pretty straight forward. Each combo variable can be aggregated into the next combo variable when producing forecasts at higher aggregations of the data. This structure resembles a pyramid, with the bottom being the lowest granularity of time series and the top being a single time series of the grand total of the data. Below is a good example with geographical combo variables that can be aggregated into one another when building a standard hierarchy forecast.

```{r, echo = FALSE, message = TRUE}
suppressMessages(library(dplyr))

message("Standard Hierarchical Time Series Data")

hts <- tibble(
  Continent = c("North America", "North America", "North America", "North America", "North America", "North America", "North America", "North America", "North America"),
  Country = c("United States", "United States", "United States", "United States", "United States", "United States", "Mexico", "Mexico", "Mexico"), 
  City = c("Kansas City", "Kansas City", "Kansas City", "Seattle", "Seattle", "Seattle", "Mexico City", "Mexico City", "Mexico City"), 
  Date = c("2020-01-01", "2020-02-01", "2020-03-01", "2020-01-01", "2020-02-01", "2020-03-01", "2020-01-01", "2020-02-01", "2020-03-01"), 
  Target = c(100, 250, 320, 80, 200, 270, 50, 80, 120)
) %>%
  dplyr::mutate(Date = as.Date(Date))

print(hts)

```

In the above example, "City" was the lowest level of the hierarchy, which feeds into "Country", which then feeds into "Continent". Finn will take this data and will forecast by City, total Country, and total Continent. After each model is ran for every level in the hierarchy, the best model is chosen at each level, then the "Best Model" and every other model is reconciled back down to the lowest level.

### Grouped Hierarchy

Grouped hierarchies are very different than the traditional hierarchy approach described above. There are some data sets that can be aggregated in various ways, meaning they need to follow another approach the hts package calls "grouped". A good example is a data set that contains historical time series by geography, customer segment, and product.

```{r, echo = FALSE, message = TRUE}
suppressMessages(library(dplyr))

message("Grouped Hierarchical Time Series Data")

gts <- tibble(
  Country = c("United States", "United States", "United States", "United States", "United States", "United States", "Mexico", "Mexico", "Mexico", "Mexico", "Mexico", "Mexico"), 
  Segment = c("Enterprise", "Enterprise", "Enterprise", "Public Sector", "Public Sector", "Public Sector", "Enterprise", "Enterprise", "Enterprise", "Enterprise", "Enterprise", "Enterprise"), 
  Product = c("Coffee", "Coffee", "Coffee", "Coffee", "Coffee", "Coffee", "Coffee", "Coffee", "Coffee", "Tea", "Tea", "Tea"), 
  Date = c("2020-01-01", "2020-02-01", "2020-03-01", "2020-01-01", "2020-02-01", "2020-03-01", "2020-01-01", "2020-02-01", "2020-03-01", "2020-01-01", "2020-02-01", "2020-03-01"), 
  Target = c(10, 20, 30, 5, 8, 11, 20, 23, 27, 50, 55, 60)
) %>%
  dplyr::mutate(Date = as.Date(Date))

print(gts)

```

It would be hard to aggregate the above data in a traditional hierarchy. The same products are found in different segments and countries, also the same segments are found in multiple countries. Finn will follow a similar modeling process as the one described for a traditional hierarchy, but instead will create forecasts at the below levels.

-   Grand Total: model "Target" across the sum of "Country", "Segment", and "Product"
-   Country: model "Target" across the sum of "Country". Creating forecasts for the grand total of "United States" and "Mexico"
-   Segment: model "Target" across the sum of "Segment". Creating forecasts for the grand total of "Enterprise" and "Public Sector"
-   Product: model "Target" across the sum of "Product". Creating forecasts for the grand total of "Coffee" and "Tea".

### External Regressors

Currently finnts does not allow for external regressors to be used in a hierarchical forecast. This is something we plan to implement in a future release. If you select a hierarchical forecast method and include external regressor input values, Finn will ignore the external regressors and will not use them in modeling.

### Spark Parallel Processing

There is also a small limitation when doing hierarchical forecasting using spark as the parallel computing back-end. The hts package finnts uses cannot handle spark data frames. So behind the scenes finnts has to bring all of the data into memory on the spark cluster driver node in order to create the hierarchies. Please keep that in mind. If you have a lot of historical data that is greater than the RAM on the spark driver node, please consider using a different VM size with more memory. The same issue holds true when reconciling the hierarchical forecast, where all of the forecasts across all time series for a specific model is loaded into the same executor node to be reconciled. Because of these limitations, we are now exploring other options outside of the hts package, including the building of our own hierarchical forecasting package. Stay tuned.
